---
title: "Exploratory Analysis of NOAA Storm Database"
author: "Grace Pehl, PhD"
date: "August 7, 2015"
output: html_document
---
# Title
Your document should have a title that briefly summarizes your data analysis

## Synopsis
Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing
### Loading the Data
```{r, message=FALSE}
library(dplyr)
```

```{r, cache=TRUE}
if (!exists("storm")) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    datafile <- "FStormData.csv.bz2"
    if (!file.exists(datafile)) {
        download.file(url, datafile, mode = "wb")
    }
    storm <- read.csv(datafile, stringsAsFactors = FALSE)
    storm <- select(storm, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
                    CROPDMG, CROPDMGEXP)
}
```
### Preprocessing
Damages to crops and property are given a code that indicates measurement units.
K - thousands, M - millions, B - billions

```{r}
storm$PROPDMG[storm$PROPDMGEXP == "K"] <- 10 ^ 3 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "K"]
storm$PROPDMG[storm$PROPDMGEXP == "M"] <- 10 ^ 6 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "M"]
storm$PROPDMG[storm$PROPDMGEXP == "B"] <- 10 ^ 9 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "B"]
storm$CROPDMG[storm$CROPDMGEXP == "K"] <- 10 ^ 3 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "K"]
storm$CROPDMG[storm$CROPDMGEXP == "M"] <- 10 ^ 6 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "M"]
storm$CROPDMG[storm$CROPDMGEXP == "B"] <- 10 ^ 9 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "B"]
```

There are 488 types of events that caused loss of life, injuries, crop damage,
or property damage.  Similarly named events will be grouped together.



```{r}
storm2 <- storm
# 14 tornado events, rank 1 by far 62.4%
storm2$EVTYPE[grepl("tornado|waterspout", tolower(storm2$EVTYPE))] <- "tornado"
# 15 hurricane events, rank 8
storm2$EVTYPE[grepl("hurricane|typhoon|tropical storm", 
                    tolower(storm2$EVTYPE))] <- "hurricane"
# 7 fire events, rank 9
storm2$EVTYPE[grepl("fire", tolower(storm2$EVTYPE))] <- "wildfire"
# 15 flash flood events, rank 7
storm2$EVTYPE[grepl("flash flood", tolower(storm2$EVTYPE))] <- "flash flood"
# 9 heat events, rank 3
 storm2$EVTYPE[grepl("heat", tolower(storm2$EVTYPE))] <- "heat"
# 131 wind events, rank 2
storm2$EVTYPE[grepl("wind", tolower(storm2$EVTYPE))] <- "wind"
# 71 events, rank 5
storm2$EVTYPE[grepl("winter|snow|ice|blizzard", 
                    tolower(storm2$EVTYPE))] <- "winter storm"
# top 10 events accout for 96.4% of deaths and injuries
# top 5 are 87.5%
# top 3 are 78.6%
# tornado, wind, heat, flood, winter storm, lightning, flash flood, hurricane, wildfire, hail
# 3 events, rank 11
storm2$EVTYPE[grepl("fog", tolower(storm2$EVTYPE))] <- "fog"
# 2, rank 12
storm2$EVTYPE[grepl("rip current", tolower(storm2$EVTYPE))] <- "rip current"
# 17, rank 13
storm2$EVTYPE[grepl("surf|seas", tolower(storm2$EVTYPE))] <- "high seas"

```
```{r}
events <- storm %>% 
          filter(PROPDMG > 0 | CROPDMG > 0 | FATALITIES > 0 | INJURIES > 0) %>% 
          group_by(EVTYPE) %>% 
          summarize(deaths = sum(FATALITIES), injuries = sum(INJURIES)) %>% 
          mutate(total = deaths + injuries) %>%
          arrange(desc(total))
dim(events)
```

```{r}
health <- storm %>% select(EVTYPE, FATALITIES, INJURIES) %>%
    filter(FATALITIES > 0 | INJURIES > 0) %>%
    group_by(EVTYPE) %>% 
    summarize(deaths = sum(FATALITIES), injuries = sum(INJURIES)) %>% 
    mutate(total = deaths + injuries) %>%
    arrange(desc(total))
# Estimates should be rounded to
# three significant digits, followed by an alphabetical character signifying the 
# magnitude of the number, i.e., 1.55B for $1,550,000,000. 
costs <- storm %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

