---
title: "US Faces Greatest Weather Risk from Tornadoes and Flooding"
author: "Grace Pehl, PhD"
date: "August 7, 2015"
output: html_document
---
## Synopsis
During the period 1950-2011, tornadoes were responsible for 62.4% of 
weather-related deaths and injuries, followed by wind, excessive heat, floods, 
and snow.  Flooding produced the greatest economic impact through floods, 
storm surge, and flash floods.  Significant damage was also 
produced by hurricanes and tornados.  

## Data Processing
### Loading the Data
```{r, message=FALSE}
library(dplyr)
```

```{r, cache=TRUE}
if (!exists("storm")) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    datafile <- "FStormData.csv.bz2"
    if (!file.exists(datafile)) {
        download.file(url, datafile, mode = "wb")
    }
    storm <- read.csv(datafile, stringsAsFactors = FALSE)
    storm <- storm %>% 
             select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
                    CROPDMG, CROPDMGEXP) %>%
             filter(PROPDMG > 0 | CROPDMG > 0 | FATALITIES > 0 | INJURIES > 0)
}
```
### Preprocessing
Damages to crops and property are given a code that indicates measurement units.
K - thousands, M - millions, B - billions

```{r}
# Multiply Out Property Values
storm$PROPDMG[storm$PROPDMGEXP == "K"] <- 10 ^ 3 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "K"]
storm$PROPDMG[storm$PROPDMGEXP == "M"] <- 10 ^ 6 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "M"]
storm$PROPDMG[storm$PROPDMGEXP == "B"] <- 10 ^ 9 * 
                                          storm$PROPDMG[storm$PROPDMGEXP == "B"]
# Multiply Out Crop Values
storm$CROPDMG[storm$CROPDMGEXP == "K"] <- 10 ^ 3 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "K"]
storm$CROPDMG[storm$CROPDMGEXP == "M"] <- 10 ^ 6 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "M"]
storm$CROPDMG[storm$CROPDMGEXP == "B"] <- 10 ^ 9 * 
                                          storm$CROPDMG[storm$CROPDMGEXP == "B"]
```

There are 488 types of events that caused loss of life, injuries, crop damage,
or property damage.  Similarly named events will be grouped together.

```{r}
storm$EVTYPE <- tolower(storm$EVTYPE)
storm$EVTYPE[grepl("tornado|waterspout", storm$EVTYPE)] <- "tornado"
storm$EVTYPE[grepl("hurricane|typhoon|tropical storm", 
                   storm$EVTYPE)] <- "hurricane"
storm$EVTYPE[grepl("fire", storm$EVTYPE)] <- "wildfire"
storm$EVTYPE[grepl("flash flood", storm$EVTYPE)] <- "flash flood"
storm$EVTYPE[grepl("heat", storm$EVTYPE)] <- "heat"
storm$EVTYPE[grepl("wind", storm$EVTYPE)] <- "wind"
storm$EVTYPE[grepl("winter|snow|ice|blizzard|wintry", storm$EVTYPE)] <- "snow"
```

## Results
### Effects of Weather Events on Population Health
```{r}
health <- storm %>% 
          select(Event = EVTYPE, FATALITIES, INJURIES) %>%
          group_by(Event) %>% 
          summarize(Deaths = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
          mutate(Event = factor(Event), Total = Deaths + Injuries) %>%
          arrange(desc(Total))
barplot(t(health[1:5, 2:3]), names.arg = health$Event[1:5], 
        legend.text = TRUE, col = c("wheat4", "wheat1"), ylab = "Individuals",
        main = "Deaths and Injuries from Weather Events 1950-2011")
```

Historically, tornadoes are responsible for 
`r round(health$Total[1]/sum(health$Total)*100, digits = 1)`% of the 
weather-related deaths and injuries in the US.  

### Economic Consequences of Weather Events
```{r}
costs <- storm %>% 
         select(Event = EVTYPE, PROPDMG, CROPDMG) %>% 
         group_by(Event) %>% 
         summarize(Property = round(sum(PROPDMG) / 10^9, digits = 1), 
                   Crops = round(sum(CROPDMG) / 10^9, digits = 1)) %>% 
         mutate(Total = Property + Crops) %>%
         arrange(desc(Total))
barplot(costs$Total[1:5], names.arg = costs$Event[1:5], col = "wheat1",
        ylab = "Damages (billions of dollars)", 
        main = "Cost of Damage to Property and Crops 
        from Weather Events 1950-2011")
```

The greatest amount of property and crop damage is produced by flooding, though
hurricanes and tornadoes also produce significant damage.  Note
that the values used have not been adjusted for inflation.
